---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# cluequery

The goal of cluequery is to provide an easy R interface to query Connectivity Map
data from [Clue](https://clue.io).

## Installation

You can install the current version of cluequery from
[Github](https://github.com/labsyspharm/cluequery) with:

```{r installation, eval=FALSE}
if (!requireNamespace("remotes", quietly = TRUE))
  install.packages("remotes")
remotes::install_github("labsyspharm/cluequery")
```

## API key

In order to use this package an API key from Clue is required. For academice
purposes, they are freely available at [Clue](https://clue.io).

The API key can be automatically retrieved by all `cluequery` functions if it is
added to the `~/.Renviron` file:

```
CLUE_API_KEY=xxx
```

## Example

This is a basic example how to query Clue for a gene signature of interest. Gene
signatures can come from any source, but must contain a set of upregulated genes
and a set of downregulated genes.

The gene signatures must be converted into
[GMT format](https://software.broadinstitute.org/cancer/software/genepattern/file-formats-guide#GMT)
for use with Clue.

### Gene signature derived from differential expression with DESeq2

In this example, we generate a gene set compatible with Clue from a
[DESeq2](https://bioconductor.org/packages/release/bioc/html/DESeq2.html)
differential expression experiment.

First, we load our example DESeq2 result:

```{r load_deseq2}
library(cluequery)
deseq2_res_path <- system.file("extdata", "example_deseq2_result.csv.xz", package = "cluequery")
deseq2_res <- read.csv(deseq2_res_path)
head(deseq2_res)
```

Note that gene IDs need to be [Entrez](https://www.ncbi.nlm.nih.gov/gene/) IDs.

We need to choose a name for the gene set and decide on a
[significance cutoff (alpha)](https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#p-values-and-adjusted-p-values)
for our differentially expressed genes.

```{r prepare_deseq2}
deseq2_gmt <- clue_gmt_from_deseq2(deseq2_res, name = "treatment_drug_x", alpha = 0.05)
str(deseq2_gmt)
```

A number of warnings are raised, indicating that some gene IDs are not part of
the [BING gene space](https://clue.io/connectopedia/category/Concept). These
genes are removed from the gene sets and not considered by Clue.

`clue_prepare_deseq2` returns a named vector with paths to the GMT
files for the up- and the down-regulated genes.

### Gene signature derived from pre-existing lists

We can also convert an pre-existing set of genes from any source to GMT files:

```{r pre_defined}
up_genes <- c(
  "10365", "1831", "9314", "4846", "678", "22992", "3397", "26136", 
  "79637", "5551", "7056", "79888", "1032", "51278", "64866", "29775", 
  "994", "51696", "81839", "23580", "219654", "57178", "7014", 
  "57513", "51599", "55818", "4005", "4130", "4851", "2050", "50650", 
  "9469", "54438", "3628", "54922", "3691", "65981", "54820", "2261", 
  "2591", "7133", "162427", "10912", "8581", "2523", "25807", "9922", 
  "30850", "4862", "8567", "79686", "55615", "51283", "3337", "2887", 
  "3223", "6915", "6907", "26056", "259217", "6574", "23097", "5164", 
  "57493", "7071", "5450", "113146", "8650"
)
down_genes <- c(
  "5128", "5046", "956", "10426", "9188", "23403", "7204", "1827", 
  "3491", "9076", "330", "8540", "22800", "10687", "19", "63875", 
  "10979", "51154", "10370", "50628", "7128", "6617", "7187", "22916", 
  "81034", "58516", "3096", "4794", "5202", "26511", "8767", "2355", 
  "22943", "1490", "133", "11010", "51025", "23160", "56902", "3981", 
  "5209", "6347", "5806", "7357", "9425", "3399", "6446", "64328", 
  "6722", "8545", "688", "861", "390", "23034", "51330", "51474", 
  "2633", "4609"
)

pre_gmt <- clue_gmt_from_list(up_genes, down_genes, "my_favourite_genes")
str(pre_gmt)
```

### Querying Clue

Now that we have GMT files, we can query Clue. Here we use the GMT files derived
from the DESeq2 result, but we could also use the `pre_gmt` files generated
above.

```{r query_clue}
submission_result <- clue_query_submit(
  deseq2_gmt[["up"]], deseq2_gmt[["down"]], name = "deseq2_query_job", use_fast_tool = FALSE
)
submission_result$result$job_id
```

`clue_query_submit` returns a nested list containing information about the submitted
job, including the job id.

Now we have to wait for the job to finish. We can use the function `clue_query_wait`
to pause our script until the results are ready.

```{r wait_result}
clue_query_wait(submission_result, interval = 60, timeout = 600)
```

`clue_query_wait` will automatically continue execution of the script once results
are ready. Now we can download and parse the results:

```{r download_and_parse}
result_path <- clue_query_download(submission_result)
result_df <- clue_parse_result(result_path)
```

```{r echo = FALSE}
knitr::kable(head(result_df, n = 10))
```

## Funding

We gratefully acknowledge support by NIH Grant 1U54CA225088-01: Systems
Pharmacology of Therapeutic and Adverse Responses to Immune Checkpoint and Small
Molecule Drugs.
